Nos\'e-Hoover Chains Improved by Backward Error Analysis: Application to Rigid-Body Molecular Dynamics

### 2018-02-12 rigid_bodies.tex
Organization of theory section

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\shadow}[1]{\widetilde{#1}}

@@

The ordinary differential equation (ODE) system that describes a Hamiltonian rigid-body motion are \cite{Silveira_2017}
\begin{subequations}
	\label{eq:ODE system for NVE}
	\begin{align}
	&\dot{\vt r} =
	%+\diff{\Ham{}}{\vt p} =
	{\mt M}^{-1} {\vt p}, \\
	&\dot{\vt p} =
	%-\diff{\Ham{}}{\vt r} = 
	{\vt F}, \\
	&\dot{\vt q} =
	%+\diff{\Ham{}}{\vt \pi} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
	&\dot{\vt \pi} =
	%-\diff{\Ham{}}{\vt q} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
	\end{align}
\end{subequations}
with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$. In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines the body's orientation, and $\vt \pi$ is the momentum conjugated to such quaternion. Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity. Matrices $\mt M$ and $\mt I$ are diagonal ones, with the former containing the body's mass in all its diagonal entries and the latter containing the principal moments of inertia. Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

@@

Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble. For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement. To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$. The equations of motion, in this case, are
\begin{subequations}
	\label{eq:ODE system for NVT}
	\begin{align}
	&\dot{\vt r} =
	{\mt M}^{-1} {\vt p}, \\
	&\dot{\vt p} =
	{\vt F} - \alpha_1 {\vt p}, \\
	&\dot{\vt q} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \\
	&\dot{\vt \pi} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
	&\dot{\eta}_j = \alpha_j, \text{ and} \\
	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M,
	\end{align}
\end{subequations}

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where each $Q_j$ is an inertial parameter of the corresponding thermostat. Besides, each $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1,
\end{align}
\end{subequations}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set as null, as usual (otherwise, $L = 6N$).\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
%Finally, $\shadow K$ is the total kinetic energy of the system, whose computation (not using Eq.~\ref{eq:kinetic energy}, hence the distinct symbol) is the central feature of the method we propose.

The ``extended energy'' which is conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H_\text{\tiny NVT} = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

As in the previous case, it is possible to devise a splitting method to solve Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}. By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation*}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation*}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain. Of course, the NVE part can be integrated exactly as described before. The NHC part, in turn, can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ enacts the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of the thermostat coordinate $\eta_1$. Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\shadow K$ to be scaled by the square of it, hence ${\shadow K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\shadow K}^\ast$. In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\shadow K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_1}$ to $e^{\frac{\timestep}{4n} \mathcal{L}_M}$ are applied again, but now in the increasing order of their indices. It is worth noting that the $\eta$'s are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H_\text{\tiny NVT}$.

The whole procedure described above is repeated $n$ times. Because of its low computational cost, the compensation required for increasing the time step $\timestep$ without degrading the accuracy of the NHC part (that is, employing a larger $n$) causes a small impact in the overall integration effort. In contrast, evaluating the NVE part is expensive for involving force/torque computations and its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}. A high-order integration scheme\cite{Omelyan_2007, vanZon_2008} could possibly admit larger time steps, but it would raise both cost and complexity because for entailing the computation (or numerical estimation) of force/torque gradients. The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development.

Here we take a simpler path. Instead of attempting to improve the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we try to quantify the obtained deviations and then use them to tune the thermostatting procedure and properly weight the sampled configurations when canonical averages are computed.

A known property of splitting methods applied to Hamiltonian ODE systems is that they provide approximate solutions which are also Hamiltonian. Thus, a trajectory generated to be roughly consistent with the specified Hamiltonian $\Ham{}$ will be exactly consistent with a nearby, albeit unknown, function $\shadow{\mathcal H}$. It is also known that $\shadow{\mathcal H}$, referred to as a shadow Hamiltonian, depends on the time step size $\timestep$ and that, for a time-reversible splitting like Eq.~\eqref{eq:splitting NVE}, $\shadow{\mathcal H} = \Ham{} + \mathcal{O}(\timestep^2)$. Estimating the trailing term of this equality, either analytically or numerically, is a task of backward error analysis. As demonstrated in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be better approximated as $\shadow{\Ham{}} = \shadow K + \shadow U + \mathcal{O}(h^4)$, where $\shadow K$ and $\shadow U$ will be called shadow kinetic and shadow potential energies, respectively. The former is given by
\begin{equation}
\label{eq:shadow kinetic energy}
\shadow K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\shadow{\mathbf \Omega}$ is a complicated matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$, but independent of both $\vt p$ and $\vt \pi$. The shadow potential is also independent of these momenta and is given by
\begin{equation}
\label{eq:shadow potential energy}
\shadow U = U - \frac{\timestep^2}{24} \left[ \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau} \tr{\mt A}(\vt q) {\mt I}^{-1} {\mt A}(\vt q) {\vt \tau} \right].
\end{equation}

Therefore, the outcome of solving Eq.~\eqref{eq:ODE system for NVE} using a Trotter-Suzuki splitting will be close to the exact solution of Eq.~\eqref{eq:ODE system for NVE}, but it will be even closer to the solution of a system composed of equations
\begin{equation}
\label{eq:shadow ODE system}
\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] =
\left[\begin{array}{c} \diff{\shadow K}{\vt p} \\ \diff{\shadow K}{\vt \pi} \end{array}\right] =
\shadow{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
as well as $\dot{\vt p} = -\diff{\shadow{\mathcal H}}{\vt r}$ and $\dot{\vt \pi} = -\diff{\shadow{\mathcal H}}{\vt q}$. In our proposed method, we acknowledge this fact and thus use the Nos\'e-Hoover chain to provide a canonical distribution consistent with $\shadow{\mathcal H}$ (rather than $\Ham{}$) at temperature $T$. Note that this entails solving Eq.~\eqref{eq:ODE system for NVT} exactly as described before, with the only requirement of replacing $K$ by $\shadow K$ in the definition of the generalized force $G_1$. By contrasting Eqs.~\eqref{eq:generalized force NVT}, \eqref{eq:shadow kinetic energy}, and \eqref{eq:shadow ODE system}, we can see that
\begin{equation}
G_1 = 2 \shadow K - L k_\text{\tiny B} T.
\end{equation}

Direct evaluation of Eq.~\eqref{eq:shadow kinetic energy} for computing $\shadow K$ is quite complex and expensive. However, one can do it very easily using the approach devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the derivatives in the left-hand side of Eq.~\eqref{eq:shadow ODE system}. Denoting the obtained estimators as $\dot{\vt r}_\text{\tiny NVE}$ and $\dot{\vt q}_\text{\tiny NVE}$ and substituting Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:shadow kinetic energy}, we find out that
\begin{equation}
\shadow K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}_\text{\tiny NVE} + \tr{\vt \pi} \dot{\vt q}_\text{\tiny NVE} \right).
\end{equation}

@@

- Scaling of $\shadow K$ during the NHC integration.

Reweighting:
\begin{equation}
\label{eq:reweighting}
\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\shadow{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
\end{equation}

@@

\subsection{Microcanonical Ensemble}

When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of a center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t) = \tr{[\begin{array}{cccc} q_1 & q_2 & q_3 & q_4 \end{array}]}$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}. In this way, the position of any constituent particle $j$ will vary with time as $\vt R_j(t) = \vt r(t) + \tr{\mt A}(\vt q(t))\vt d_j$, where $\vt d_j$ is the (constant) position of particle $j$ in a body-fixed frame of reference centered in $\vt r$ and ${\mt A}(\vt q)$ is the body's orientation matrix (that is, the rotation matrix connecting the body-fixed and space-fixed frames) expressed as a function of the quaternion components. The superscript $t$ stands for transposition, which is equivalent to inversion in the case of a canonical transformation like $\mt A$ \cite{Goldstein_2002}. The central feature explored in Ref.~\citenum{Silveira_2017} is the fact that $\mt A$ admits the simple factorization
\begin{equation}
\label{eq:factorization_of_A}
{\mt A}(\vt q) = \tr{\mt B}(\vt q) {\mt C}(\vt q),
\end{equation}
where
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

>>>>

\newcommand{\mt}[1]{\boldsymbol{\mathbf{#1}}}   % matrix symbol
\newcommand{\vt}[1]{\boldsymbol{\mathbf{#1}}}   % vector symbol
\newcommand{\tr}[1]{#1^\text{t}}                % transposition
\newcommand{\diff}[2]{\frac{\partial #1}{\partial #2}} % derivative
%\newcommand{\diff}[2]{\nabla_{#2}#1}            % derivative
\newcommand{\sdiff}[2]{\nabla^2_{#2}#1}         % 2nd derivative
\newcommand{\Ham}[1]{{\mathcal H}_\text{#1}}    % Hamiltonian
\newcommand{\Liu}[1]{i\!L_\text{#1}}            % Liouville
\newcommand{\timestep}{h}
\newcommand{\modified}[1]{\widetilde{#1}}

@@

The ordinary differential equation (ODE) system that describes a Hamiltonian rigid-body motion are \cite{Silveira_2017}
\begin{subequations}
	\label{eq:ODE system for NVE}
	\begin{align}
%
	&\dot{\vt r} =
	%+\diff{\Ham{}}{\vt p} =
	{\mt M}^{-1} {\vt p}, \\
%
	&\dot{\vt p} =
	%-\diff{\Ham{}}{\vt r} = 
	{\vt F}, \\
%
	&\dot{\vt q} =
	%+\diff{\Ham{}}{\vt \pi} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \label{eq:EDO_q} \\
%
	&\dot{\vt \pi} =
	%-\diff{\Ham{}}{\vt q} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau, \label{eq:EDO_pi}
	\end{align}
\end{subequations}
with $\vt \omega = \frac{1}{2} {\mt I}^{-1} \tr{\mt B}(\vt q) {\vt \pi}$. In these equations, $\vt r$ is the center-of-mass position of the body, $\vt p$ is its linear momentum, $\vt q$ is a unit quaternion that determines the body's orientation, and $\vt \pi$ is the momentum conjugated to such quaternion. Vectors $\vt F$ and $\vt \tau$ are, respectively, the resultant force and resultant torque exerted on the body, both represented in the space-fixed frame of reference, while $\vt \omega$ is the three-dimensional angular velocity. Matrices $\mt M$ and $\mt I$ are diagonal ones, with the former containing the body's mass in all its diagonal entries and the latter containing the principal moments of inertia. Finally, the matrix-valued functions $\mt B$ and $\vt C$ are defined as
\begin{equation*}
\label{eq:def_B_and_C}
\mt B(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 & -q_4 &  q_3 \\
 q_4 &  q_1 & -q_2 \\
-q_3 &  q_2 &  q_1
\end{smallarray}
\; \text{and} \;
\mt C(\vt q) = \begin{smallarray}{rrrr}
-q_2 & -q_3 & -q_4 \\
 q_1 &  q_4 & -q_3 \\
-q_4 &  q_1 &  q_2 \\
 q_3 & -q_2 &  q_1
\end{smallarray}.
\end{equation*}

@@

Here we propose a new method for simulating a system of rigid bodies in the canonical ensemble. For this, we couple a single Nos\'{e}-Hoover thermostat chain \cite{Martyna_1992} to all translational and rotational degrees of freedom of the rigid-body system. This is simpler than the approach of Kamberaj \textit{et al}.,\cite{Kamberaj_2005} who employed separate thermostat chains to these two classes of movement. To this end, we consider an extra generalized coordinate $\eta_j$ and its conjugate momentum $p_{\eta_j}$ for each thermostat $j$, with $1 \leq j \le M$. The equations of motion, in this case, are
\begin{subequations}
	\label{eq:ODE system for NVT}
	\begin{align}
%
\label{eq:nhc_r}
	&\dot{\vt r} =
	{\mt M}^{-1} {\vt p}, \\
%
\label{eq:nhc_p} 
	&\dot{\vt p} =
	{\vt F} - \alpha_1 {\vt p},\\
%
\label{eq:nhc_q}
	&\dot{\vt q} =
	\frac{1}{2} \mt B(\vt q) \vt \omega, \text{ and} \\
%
\label{eq:nhc_pi}
	&\dot{\vt \pi} =
	\frac{1}{2} \mt B(\vt \pi) \vt \omega + 2 \mt C(\vt q) \vt \tau - \alpha_1 {\vt \pi}, \\
%
\label{eq:nhc_eta}
	&\dot{\eta}_j = \alpha_j, \text{ and} \\
%
\label{eq:nhc_p_eta}
	&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j} \qquad \text{for} \quad 1 \leq j \le M,
	\end{align}
\end{subequations}

In these equations, $\alpha_j$ $=$ ${p_{\eta_j}}/{Q_j}$ for $1 \leq j \le M$ and $\alpha_{M+1} = 0$, where each $Q_j$ is an inertial parameter of the corresponding thermostat. Besides, each $G_j$ is a generalized force acting on thermostat $j$, defined as
\begin{subequations}
\begin{align}
\label{eq:generalized force NVT}
&G_1 = \tr{\vt p} \diff{K}{\vt p} + \tr{\vt \pi} \diff{K}{\vt \pi} - L k_\text{\tiny B} T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_\text{\tiny B} T  \qquad \text{for} \quad j > 1,
\end{align}
\end{subequations}
where $k_\text{\tiny B}$ is the Boltzmann constant, $T$ is the system temperature, and $L = 6N-3$ if the vector-sum of all linear momenta is set as null, as usual (otherwise, $L = 6N$).\cite{Martyna_1994} As recommended in Ref.~\citenum{Martyna_1992}, one can make $Q_1 = L k_\text{\tiny B} T t_d^2$ and $Q_j = k_\text{\tiny B} T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the system.
%Finally, $\modified K$ is the total kinetic energy of the system, whose computation (not using Eq.~\eqref{eq:kinetic energy}, hence the distinct symbol) is the central feature of the method we propose.
The ``extended energy'' which is conserved by the non-Hamiltonian system in Eq.~\eqref{eq:ODE system for NVT} is \cite{Martyna_1992}
\begin{equation}
\label{eq:nvt extended energy}
H = \Ham{} + {\textstyle\sum\limits_{j=1}^{M}} \frac{p_{\eta_j}^2}{2Q_j} + L k_\text{\tiny B} T\eta_1 + k_\text{\tiny B} T {\textstyle\sum\limits_{j=2}^{M}} \eta_j.
\end{equation}

As in the previous case, it is possible to devise a splitting method to solve Eq.~\eqref{eq:ODE system for NVT} numerically \cite{Tuckerman_2010}. By defining a propagator $e^{\timestep \mathcal{L}_\text{\tiny NVT}}$, where $\mathcal{L}_\text{\tiny NVT}$ is a generalized (i.e. non-Hamiltonian) Liouville operator, the splitting goes as
\begin{equation}
\label{eq:splitting NVT}
e^{\timestep \mathcal{L}_\text{\tiny NVT}} = e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} e^{\timestep \Liu{\tiny NVE}} e^{\frac{\timestep}{2} \mathcal{L}_\text{\tiny NHC}} + \mathcal{O}(\timestep^3),
\end{equation}
where $\Liu{\tiny NVE}$ is the same operator of the previous case and $\mathcal{L}_\text{\tiny NHC}$ aggregates all new terms introduced by the Nos\'e-Hoover chain. Of course, the NVE part can be integrated exactly as described before. The NHC part, in turn, can be split even further as
\begin{equation*}
e^{\frac{\timestep}{2} \mathcal{L}_\text{NHC}} = \left[ \left( \textstyle\prod\limits_{j=M}^1 e^{\frac{\timestep}{4n} \mathcal{L}_j }\right) e^{\frac{\timestep}{2n} \mathcal{L}_0 } \left(  \textstyle\prod\limits_{j=1}^M e^{\frac{\timestep}{4n} \mathcal{L}_j }\right)  \right]^n.
\end{equation*}

@@

After that, propagator $e^{\frac{\timestep}{2n} \mathcal{L}_0}$ enacts the effect of thermostat $1$ on the motion of the rigid bodies, as well as the evolution of the thermostat coordinate $\eta_1$. Its action is expressed as ${\vt p} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt p}^\ast$, ${\vt \pi} = e^{-\alpha_1^\ast \frac{\timestep}{2n}} {\vt \pi}^\ast$, and $\eta_1 = \eta_1^\ast + \alpha_1^\ast \frac{\timestep}{2n}$.
%Note that momentum vectors $\vt p$ and $\vt \pi$ are scaled the same factor, which causes $\modified K$ to be scaled by the square of it, hence ${\modified K} = e^{-\alpha_1^\ast \frac{\timestep}{n}} {\modified K}^\ast$. In Sec.~\ref{sec:backward error analysis} we demonstrate why such effect on $\modified K$ is observed even though we do not use Eq.~\eqref{eq:kinetic energy} to compute it.
Finally, propagators $e^{\frac{\timestep}{4n} \mathcal{L}_1}$ to $e^{\frac{\timestep}{4n} \mathcal{L}_M}$ are applied again, but now in the increasing order of their indices. It is worth noting that the $\eta$'s are driven variables with no influence on the dynamics, but they are often computed just for checking how well the integrator conserves $H$.

The whole procedure described above is repeated $n$ times. Because of its low computational cost, the compensation required for increasing the time step $\timestep$ without degrading the accuracy of the NHC part (i.e. employing a larger $n$) causes a small impact in the overall integration effort. In contrast, evaluating the NVE part is expensive for involving force/torque computations and its accuracy goes down very quickly as $\timestep$ increases \cite{Davidchack_2010, Silveira_2017}. A high-order integration scheme\cite{Omelyan_2007, vanZon_2008} could possibly admit larger time steps, but it would raise both cost and complexity for entailing the computation (or numerical estimation) of force/torque gradients. The processed splitting approach \cite{Omelyan_2008} seems promising, but its extension to the NVT case without doing back-and-forth processing at every time step would require further theoretical development. Nevertheless, we choose to take a simpler path. Instead of attempting to improve the evaluation of $e^{\timestep \Liu{\tiny NVE}}$, we try to quantify the obtained deviations and then use them to tune the thermostatting procedure and properly weight the sampled configurations when canonical averages are computed.

A known property of splitting methods applied to Hamiltonian ODE systems is that they provide Hamiltonian approximate solutions. In other words, a trajectory generated to be roughly consistent with the specified function $\Ham{}$ will be exactly consistent with a nearby (albeit unknown) function $\Ham{S}$, referred to as a shadow Hamiltonian. Thus, whereas $\Ham{}$ exhibits a certain degree of oscillation along that trajectory, the value of $\Ham{S}$ is actually conserved (up to round-off errors). It is also known that $\Ham{S}$ depends on the time step size $\timestep$ and that, for a time-reversible splitting such as Eq.~\eqref{eq:splitting NVE}, one can write $\Ham{S} = \Ham{} + \mathcal{O}(\timestep^2)$. Estimating the trailing term is a task of backward error analysis.

As shown in Sec.~\ref{sec:backward error analysis}, the shadow Hamiltonian for a Verlet-type method applied to a system of rigid bodies can be better approximated by $\Ham{S} = \modified K + \modified U + \mathcal{O}(h^4)$, where $\modified K$ and $\modified U$ are named modified kinetic and modified potential energies, respectively. The former can be expressed as
\begin{equation}
\label{eq:shadow kinetic energy}
\modified K = \frac{1}{2} \tr{ \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right]} \modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
where $\modified{\mathbf \Omega}$ is a symmetric matrix-valued function of $\vt r$, $\vt q$, and $\timestep^2$. The modified potential, which is also independent of $\vt p$ and $\vt \pi$, is given by
\begin{equation}
\label{eq:shadow potential energy}
\modified U = U - \frac{\timestep^2}{24} \left[ \tr{\vt F} {\mt M}^{-1} {\vt F} + \tr{\vt \tau}_b {\mt I}^{-1} {\vt \tau}_b \right],
\end{equation}
where ${\vt \tau}_b = {\mt A}(\vt q) {\vt \tau}$ is the body-fixed representation of the torques. Therefore, the outcome of solving Eq.~\eqref{eq:ODE system for NVE} using a Trotter-Suzuki splitting will be close to the exact solution of Eq.~\eqref{eq:ODE system for NVE}, but it will be even closer to the solution of
\begin{equation}
\label{eq:shadow ODE system}
\left[\begin{array}{c} \dot{\vt r} \\ \dot{\vt q} \end{array}\right] =
\left[\begin{array}{c} \diff{\modified K}{\vt p} \\ \diff{\modified K}{\vt \pi} \end{array}\right] =
\modified{\mathbf \Omega}(\vt r, \vt q, \timestep^2) \left[\begin{array}{c} \vt p \\ \vt \pi \end{array}\right],
\end{equation}
together with $\dot{\vt p} = -\diff{\modified{\Ham{}}}{\vt r}$ and $\dot{\vt \pi} = -\diff{\modified{\Ham{}}}{\vt q}$, where $\modified{\Ham{}} = \modified K + \modified U$. Our proposal consists in acknowledging this fact and then tuning the Nos\'e-Hoover chain to provide a canonical distribution consistent with $\modified{\Ham{}}$ (instead of $\Ham{}$) at temperature $T$. Fortunately, this entails solving Eq.~\eqref{eq:ODE system for NVT} almost exactly as described before. The only requirement is to replace $K$ by $\modified K$ in the definition of the generalized force in thermostat $1$, given by Eq.~\eqref{eq:generalized force NVT}. By contrasting this one with Eqs.~\eqref{eq:shadow kinetic energy} and \eqref{eq:shadow ODE system}, we observe that such force becomes
\begin{equation}
G_1 = 2 \modified K - L k_\text{\tiny B} T.
\end{equation}

It is possible to evaluate Eq.~\eqref{eq:shadow kinetic energy} directly for computing $\modified K$, but this is quite complex and expensive. Instead, one can do it very easily by means of a method devised by \citeauthor{Eastwood_2010} \cite{Eastwood_2010}, which consists in using numerical differentiation to estimate the derivatives in the left-hand side of Eq.~\eqref{eq:shadow ODE system}. By denoting the obtained estimators as $\dot{\vt r}_\text{\tiny NVE}$ and $\dot{\vt q}_\text{\tiny NVE}$, and substituting Eq.~\eqref{eq:shadow ODE system} into Eq.~\eqref{eq:shadow kinetic energy}, we find out that
\begin{equation}
\modified K = \frac{1}{2} \left( \tr{\vt p} \dot{\vt r}_\text{\tiny NVE} + \tr{\vt \pi} \dot{\vt q}_\text{\tiny NVE} \right).
\end{equation}

@@

- Scaling of $\modified K$ during the NHC integration.

Reweighting:
\begin{equation}
\label{eq:reweighting}
\langle A \rangle = \frac{\sum\limits_{i=1}^{N_s} A(\vt x_i) \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right]}{\sum\limits_{i=1}^{N_s} \exp \left[{\frac{\modified{\Ham{}}(\vt x_i) - {\Ham{}}(\vt x_i)}{k_\text{\tiny B} T}}\right] }.
\end{equation}

@@

\subsection{Hamiltonian Dynamics}
\label{eq:hamiltonian dynamics}

When describing the motion of a rigid body, the translational and rotational modes can be decoupled and expressed, respectively, as the evolution of the center-of-mass coordinate $\vt r(t)$ in a space-fixed frame of reference and the evolution of a quaternion $\vt q(t)$ which continuously satisfies the unit-norm constraint $\tr{\vt q}{\vt q} = 1$ \cite{Goldstein_2002}. In this way, the connection between the body-fixed and space-fixed frames of reference can be recovered at any time by the rotation matrix $\mt A(\vt q)$, which admits the factorization mentioned in Sec.~\ref{sec:methodology}. This factorization creates an intermediate frame of reference for representing three-dimensional vectors in $\mathbb{R}^3$ as quaternions that lie in the hyperplane orthogonal to $\vt q$, since the columns of matrix $\mt B$ and those of matrix $\mt C$ form two different bases for such a subspace of $\mathbb{R}^4$. Transformations amongst three distinct reference frames are enacted by multiplications with these matrices or their transposes, such as represented in the following scheme:
\begin{equation*}
\boxed{\substack{\text{Body-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\mt B}_{\tr{\mt B}}}
\boxed{\substack{\text{Quaternion-fixed} \\ \text{frame}}}
\mathrel{\mathop{\rightleftarrows}^{\tr{\mt C}}_{\mt C}}
\boxed{\substack{\text{Space-fixed} \\ \text{frame}}}
\end{equation*}

The intermediate frame is exactly the subspace of $\mathbb{R}^4$ that must contain the rate-of-change vector $\dot{\vt q}$ if the norm of $\vt q$ is to remain constant, since $\tr{\vt q} {\vt q} = 1$ implies that $\tr{\vt q} \dot{\vt q} = 0$. In the Hamiltonian approach for the motion of a rigid body, the momentum $\vt \pi$ conjugated to $\vt q$ is given by $\vt \pi = 2 \mt B(\vt q) \mt I \vt \omega$.\cite{Silveira_2017} This means that $\vt \pi$ is the quaternion-fixed representation of the angular momentum $\vt L = \mt I \vt \omega$, thus being orthogonal to $\vt q$ as well.
